---
title: "Week 3, Part 1"
subtitle: "Spatial Predicates & Filtering"
format:
  revealjs: 
    smaller: false
    incremental: false
    logo: ../csu-rams-logo.png
    slide-number: c/t
    footer: "[ESS 523c: Environmental Data Science Applications: Water Resources](https://github.com/mikejohnson51/csu-ess-523c)"
    include-before-body: header.html
    include-after-body: footer-annotations.html
    theme: [default, csu-css.scss]
    width: 1600
    height: 1000
    countIncrementalSlides: false
    title-slide-attributes:
      data-background-color: "#1E4D2B"
knitr:
  opts_chunk: 
    echo: true
    message: true
    collapse: true
    comment: "#>"
    out.width: "100%"
---

```{r, include = F}
library(sf)
library(tidyverse)
library(knitr)
library(AOI)

states <- aoi_get(state = "all") |> filter(state_abbr %in% c("WA", "OR", "MT", "ID")) |> select(name)
wa <- filter(states, name == "Washington")

cities <- read_csv("data/uscities.csv") |>
  st_as_sf(coords = c("lng", "lat"), crs = 4326) |>
  select(city, state_name, population)
```

# Quick Recap

## Last Week: Measurements and Coordinates

You learned how to:

- **Measure** spatial data: distances, areas, lengths (GEOS library)
- **Transform** between coordinate systems (PROJ library)
- **Distinguish** between geodesic (curved Earth) and planar (flat space) calculations
- **Understand** how S2 geometry makes geodesic calculations practical

. . .

**Key insight**: Spatial libraries (GDAL, PROJ, GEOS, S2) handle the math. Your job: ask the right questions and interpret the results correctly.

---

Today we focus on **asking spatial questions** about relationships between geometries.

# Topological Relationships

## Geometry Has Structure

Every geometry has three parts:

- **Interior** — the space *inside* the geometry (0D, 1D, or 2D depending on type)
- **Boundary** — the edge or envelope that defines the limit
- **Exterior** — everything *outside*

. . .

This structure matters for answering questions like:

- Does this polygon *contain* that point?
- Does this stream *intersect* that watershed?
- Are these county boundaries *touching* with no overlap?

## Interior, Boundary, Exterior: Examples {.smaller}

:::{.columns}
:::{.column width="33%"}

**POINT (0D)**
- Interior: the point itself
- Boundary: empty
- Exterior: everything else

:::
:::{.column width="33%"}

**LINESTRING (1D)**
- Interior: all points except endpoints
- Boundary: the two endpoints
- Exterior: everything else

:::
:::{.column width="33%"}

**POLYGON (2D)**
- Interior: the area enclosed
- Boundary: the rings (edges)
- Exterior: outside the rings

:::
:::

# Spatial Predicates: The DE-9IM Model

## Asking "What's the relationship?"

When two geometries meet or overlap, we can ask: **exactly what kind of relationship exists?**

This is formalized in the **Dimensionally Extended 9-Intersection Model (DE-9IM)**, a mathematical framework that tests *all possible intersections* between two geometries.

**Key insight**: The type of overlap tells you the spatial relationship.

- Point AND point overlapping = intersection
- Line AND boundary = touching (but not overlapping internally)
- Area AND area with gap at edge = overlapping

## The DE-9IM Matrix Structure

For geometries *a* and *b*, test 9 intersections:

$$\begin{bmatrix} I_a \cap I_b & I_a \cap B_b & I_a \cap E_b \\ B_a \cap I_b & B_a \cap B_b & B_a \cap E_b \\ E_a \cap I_b & E_a \cap B_b & E_a \cap E_b \end{bmatrix}$$

Where:
- **I** = interior (the "inside")
- **B** = boundary (the edge)
- **E** = exterior (everything outside)

Each cell records what *dimension* of overlap occurs:
- **F** = False (empty, no intersection)
- **0** = Point (0D intersection)
- **1** = Line (1D intersection)
- **2** = Area (2D intersection)
- **\*** = Any dimension (wildcard in some systems)

## DE-9IM Example: Point in Polygon {.smaller}

**Is a point P inside polygon A?**

| | Interior of A | Boundary of A | Exterior of A |
|---|---|---|---|
| **Interior of P** | — | — | — |
| **Boundary of P** | 0 | F | F |
| **Exterior of P** | F | F | 2 |

Result: `0F*FFF*`

- Point boundary (itself) intersects polygon interior at dimension 0 ✓
- Point boundary does NOT intersect polygon boundary
- Point exterior intersects polygon exterior (the space outside)

**Interpretation**: Point is *inside* the polygon.

## DE-9IM Example: Overlapping Polygons {.smaller}

**Do two overlapping counties A and B share area?**

| | Interior of B | Boundary of B | Exterior of B |
|---|---|---|---|
| **Interior of A** | 2 | 1 | 2 |
| **Boundary of A** | 1 | F | 1 |
| **Exterior of A** | 2 | 1 | 2 |

Result: `212101212`

- Interiors overlap (2D) ✓
- Boundaries cross (1D) ✓
- Areas extend outside each other (2D exteriors)

**Interpretation**: Polygons *overlap* with partial area shared.

## DE-9IM Example: Touching Boundaries {.smaller}

**Do two counties share a border but no area (A touches B)?**

| | Interior of B | Boundary of B | Exterior of B |
|---|---|---|---|
| **Interior of A** | F | 1 | 2 |
| **Boundary of A** | 1 | F | 1 |
| **Exterior of A** | 2 | 1 | 2 |

Result: `F1F1*1212`

- Interiors do NOT intersect (F) — no overlapping area
- Boundaries DO intersect (1D) — they share an edge
- Each interior is in the other's exterior (F vs 2)

**Interpretation**: Polygons *touch* at a boundary; interiors stay separate.

## Simplifying to Named Predicates {.smaller}

The full matrix is complex, so we use **shorthand predicates** that check for specific patterns:

- **Intersects** — ANY intersection exists (matrix has non-F values)
- **Touches** — Boundaries (B) meet; interiors (I) don't overlap
- **Within** — Entire I_a ⊆ I_b ∪ B_b (A inside B)
- **Contains** — Entire I_b ⊆ I_a ∪ B_a (B inside A)
- **Overlaps** — Same dimension; partial area/length overlap (not fully within)
- **Crosses** — Different dimensions; lower-dim geometry crosses through higher-dim

## Why This Matters: Topological Invariance {.smaller}

A crucial property: **Topological relationships are preserved under coordinate transformations.**

If polygon A is *within* polygon B:
- True in WGS84 (geographic, curved Earth)
- Still true in EPSG:5070 (projected, flat map)
- Still true after any CRS transformation

The DE-9IM result never changes—it depends only on spatial structure, not coordinates.

::: callout-important
This is why spatial predicates are different from distance-based operations. Distances *change* under transformation; relationships don't.
:::

## Visual Guide: The Six Named Predicates {.smaller}

:::{.columns}
:::{.column width="50%"}

**Intersects** — Share any space
```
  [A]
   [B]  
   →→
```

**Touches** — Boundary meets boundary; no area overlap
```
  [A] [B]
    |
```

**Within** — A completely inside B
```
  [B:  [A]  ]
```

:::
:::{.column width="50%"}

**Contains** — B completely inside A
```
  [A:  [B]  ]
```

**Overlaps** — Same dimension, partial overlap
```
  [A]
   [B]
   →→
```

**Crosses** — Lower-dim geometry crosses higher-dim
```
  -------  (line A)
    |
  [B]   (polygon B)
```

:::
:::

## Using Predicates in R

`sf` provides functions for each predicate:

```{r}
#| eval: false
st_intersects(x, y)   # Do x and y share any space?
st_touches(x, y)       # Boundary meets boundary?
st_within(x, y)        # Is x completely inside y?
st_contains(x, y)      # Does x completely contain y?
st_overlaps(x, y)      # Partial overlap, same dimension?
st_crosses(x, y)       # Does x cross through y?
st_disjoint(x, y)      # No intersection?
```

Each returns a **sparse matrix of indices**:

```{r}
#| eval: false
st_touches(states, wa)
# [[1]] integer(0)       # Oregon doesn't touch WA
# [[2]] integer(0)       # California doesn't touch WA
# [[3]] integer(1)       # Montana touches WA at index 1
# [[4]] integer(1)       # Idaho touches WA at index 1
```

Row positions = states; returned index 1 = Washington's position.

## From Named Predicates to Full DE-9IM: `st_relate()` {.smaller}

The named predicates (intersects, touches, etc.) are **masks** over the underlying DE-9IM string code.

Use `st_relate()` to get the **full 9-character DE-9IM string**:

```{r}
#| eval: false
st_relate(states, wa)
# Returns DE-9IM codes like "2FFF1FFF2", "F1F1*1212", "212101212", etc.
```

Compare:

```{r}
#| eval: false
# Named predicates (boolean)
st_touches(states, wa, sparse = FALSE)
#      [,1]
# [1,] FALSE  # Oregon
# [2,] FALSE  # California  
# [3,]  TRUE  # Montana
# [4,]  TRUE  # Idaho

# Full DE-9IM string
st_relate(states, wa, sparse = FALSE)
#           [,1]
# [1,] "2FFF1FFF2"  # Oregon: overlaps in 2D (polygon), boundary touches
# [2,] "2FFF1FFF2"  # California: similar
# [3,] "F1F1*1212"  # Montana: touches (boundary meets boundary)
# [4,] "F1F1*1212"  # Idaho: touches
```

**Why use full codes?** For specialized applications where you need to distinguish subtle topological relationships. Most of the time, named predicates are sufficient.

## Sparse vs. Dense Results {.smaller}

**Sparse matrix** (default): Efficient; only stores non-zero indices. Fast for filtering.

**Dense matrix** (visual): Use `sparse = FALSE` for logical TRUE/FALSE. Better for reading.

```{r}
#| eval: false
st_touches(states, wa, sparse = FALSE)
#      [,1]
# [1,] FALSE  # Oregon
# [2,] FALSE  # California
# [3,]  TRUE  # Montana
# [4,]  TRUE  # Idaho
```

This logical matrix maps directly to rows: extract with `states[result == TRUE]`.

## Pattern: Match and Extract

Query tests all geometries in x against all in y. Index result to extract matches:

```{r}
#| eval: false
# Which states touch Washington? Get logical vector.
touch_matrix <- st_touches(states, wa, sparse = FALSE)[, 1]

# Extract rows where TRUE
neighbors <- states[touch_matrix, ]

# Same result, shorthand:
neighbors <- st_filter(states, wa, .predicate = st_touches)
```

`st_filter()` automates this pattern—recommended for cleaner code.



## The Power of Predicates: Filter by Space

Instead of filtering by a column value (`name == "Washington"`), filter by a *spatial relationship*:

```{r}
#| eval: false
# Traditional filter (attribute-based)
filter(states, name == "Washington")

# Spatial filter (predicate-based)
st_filter(states, wa, .predicate = st_touches)
```

Both return subsets of `states`, but spatial filter uses topology.

. . .

**Default predicate is `st_intersects`** — returns all geometries that touch or overlap the target.

```{r}
#| eval: false
st_filter(states, wa)  # = st_filter(states, wa, .predicate = st_intersects)
```

## Real Example: Which States Touch Washington?

```{r}
#| eval: false
wa_neighbors <- st_filter(states, wa, .predicate = st_touches)
```

```{r}
#| eval: true
#| echo: false
#| fig.align: center
#| fig.width: 10
#| fig.height: 6

wa_neighbors <- st_filter(states, wa, .predicate = st_touches)

ggplot(states) + 
  geom_sf(fill = "white", color = "black", linewidth = 0.5) + 
  geom_sf(data = wa, fill = "steelblue", alpha = 0.4, linewidth = 1) +
  geom_sf(data = wa_neighbors, fill = "orange", alpha = 0.6, linewidth = 0.8) + 
  theme_void() +
  labs(title = "States Touching Washington (orange)")
```

## Distance-Based Filtering

Predicates can also use distance thresholds:

```{r}
#| eval: false
# "Which cities are within 100 km of Fort Collins?"
fc <- filter(cities, city == "Fort Collins")

nearby_cities <- st_filter(
  cities, 
  fc, 
  .predicate = st_is_within_distance, 
  dist = 100000  # units in CRS (meters for geographic)
)
```

Distance units match the CRS:
- Geographic CRS (WGS84): meters
- Projected CRS (EPSG:5070): meters
- Must account for CRS when specifying distance

. . .

**Hydrology applications**:
- Which stream gauges are within 50 km upstream?
- Which reservoirs are within 100 km of this city?
- Which monitoring stations bracket this watershed outlet?

## Visual: Multiple Predicates Compared {.smaller}

```{r}
#| eval: true
#| echo: false
#| fig.align: center
#| fig.width: 14
#| fig.height: 5

# Create comparison plots
p_intersects <- ggplot(states) + 
  geom_sf(fill = "white", color = "black", linewidth = 0.3) + 
  geom_sf(data = wa, fill = "steelblue", alpha = 0.3) +
  geom_sf(data = st_filter(states, wa, .predicate = st_intersects), 
          fill = "red", alpha = 0.5) + 
  theme_void() +
  labs(title = "st_intersects()")

p_touches <- ggplot(states) + 
  geom_sf(fill = "white", color = "black", linewidth = 0.3) + 
  geom_sf(data = wa, fill = "steelblue", alpha = 0.3) +
  geom_sf(data = st_filter(states, wa, .predicate = st_touches), 
          fill = "orange", alpha = 0.5) + 
  theme_void() +
  labs(title = "st_touches()")

p_disjoint <- ggplot(states) + 
  geom_sf(fill = "white", color = "black", linewidth = 0.3) + 
  geom_sf(data = wa, fill = "steelblue", alpha = 0.3) +
  geom_sf(data = st_filter(states, wa, .predicate = st_disjoint), 
          fill = "lightgreen", alpha = 0.5) + 
  theme_void() +
  labs(title = "st_disjoint()")

gridExtra::grid.arrange(p_intersects, p_touches, p_disjoint, nrow = 1)
```

Red = intersects, Orange = touches, Green = disjoint from Washington.

## Example 1: States Touching Washington {.smaller}

```{r}
#| eval: false
# Which states share a border with Washington?
wa_neighbors <- st_filter(states, wa, .predicate = st_touches)
```

**MT and ID** return TRUE (they touch WA's boundary).
**OR** returns FALSE (touches WA, but we're using `st_touches` which requires boundary contact, not intersection).

Actually: Oregon DOES touch Washington. Let me use `st_intersects`:

```{r}
#| eval: false
wa_adjacent <- st_filter(states, wa, .predicate = st_intersects)
# Returns: Montana, Idaho, Oregon (all share boundary or overlap)
```

## Example 2: Cities Within Each State {.smaller}

**Question**: For each state, how many cities fall inside?

```{r}
#| eval: false
# Using st_within: city inside state?
cities_in_states <- st_filter(cities, states, .predicate = st_within)

# Or aggregate seamlessly with group_by:
cities_by_state <- st_join(cities, states, join = st_within) |>
  group_by(name) |>
  summarise(n_cities = n(), .groups = "drop")
```

## Spatial Joining Example {.smaller}

**Question**: Which state does each city belong to?

```{r}
#| eval: true
#| echo: false
#| fig.align: center
#| fig.width: 12
#| fig.height: 5

# Join cities to states
cities_with_state <- st_join(cities, states, join = st_within)

# Plot before (left) and after (right)
p_before <- ggplot(states) + 
  geom_sf(fill = "white", color = "navy", linewidth = 0.3) + 
  geom_sf(data = cities, color = "red", size = 1) + 
  theme_void() +
  labs(title = "Cities (red) in State Polygons")

p_after <- ggplot(cities_with_state) + 
  geom_sf(aes(color = name), size = 1.5, show.legend = FALSE) + 
  geom_sf(data = states, fill = NA, color = "navy", linewidth = 0.3) + 
  scale_color_manual(values = c("Washington" = "red", "Oregon" = "orange", 
                                "Montana" = "purple", "Idaho" = "darkgreen")) +
  theme_void() +
  labs(title = "Cities colored by state (after st_join)")

gridExtra::grid.arrange(p_before, p_after, nrow = 1)
```

After joining, cities inherit state name column and are colored accordingly.

```{r}
#| eval: false
# Which USGS stream gauges fall within each HUC8 watershed?

gauges_in_watershed <- st_join(gauges, watersheds, join = st_within)

# Summary: Average discharge per watershed
discharge_summary <- gauges_in_watershed |>
  group_by(huc8, watershed_name) |>
  summarise(
    n_gauges = n(),
    mean_discharge_m3s = mean(discharge_m3s, na.rm = T),
    .groups = "drop"
  )
```

**Key**: `st_join()` can produce many-to-many matches. If a watershed has 500 gauges on its streams, the result has 500 rows. `group_by()` + `summarise()` aggregates meaningfully.



# Spatial Joining: Combining Data by Relationship

## From Filtering to Joining

**Filtering**: Subset one dataset based on spatial relationship to another (select rows).

**Joining**: Combine columns from two datasets based on spatial relationships (connect datasets).

- **Filter** → Returns subset of x
- **Join** → Returns all of x with columns from matching y rows appended

Joins are where spatial analysis becomes truly powerful — merging data from multiple sources using space as the key.

## st_join: The Core Spatial Join Function {.smaller}

For each row in the first dataset, find matching rows in the second and add their columns:

```{r}
#| eval: false
# For each city, which state contains it?
cities_with_state <- st_join(cities, states)
```

Returns `cities` with all columns from `states` appended.

**Important**: If a city matches multiple states (shouldn't happen, but geometrically possible), you get multiple rows for that city. This is a **many-to-many join**.

## Join Type and Predicate {.smaller}

Two parameters control matching:

**Join type** (`left`, `inner`):
- `left = TRUE` (default): Keep all rows from x; add NA if no match
- `left = FALSE` (inner): Keep only rows from x that match with y

**Predicate** (`join` argument), e.g.:
```{r}
#| eval: false
# Default: st_intersects
st_join(cities, states)

# City must be completely inside state
st_join(cities, states, join = st_within)

# City boundary touches state boundary
st_join(cities, states, join = st_touches)

# State must fully contain city
st_join(cities, states, join = st_contains)
```

## The Many-to-Many Join Hazard {.smaller}

Spatial joins often produce unexpected duplication:

```{r}
#| eval: false
# 10 counties × 5 starbucks per county average
# Result: 50 rows (not 10!)
county_sb <- st_join(counties, starbucks)  

# Without grouping first, you lose information
head(county_sb)
#> # A tibble: 50 × 25 (includes all starbucks columns!)
#> # Each county repeated 5 times
```

**Solution**: Always aggregate after spatial joins:

```{r}
#| eval: false
county_starbucks <- st_join(counties, starbucks) |>
  group_by(county_id, county_name) |>
  summarise(n = n(), .groups = "drop")
```

Now you have 1 row per county with the count.

## Aggregation Pattern: After Joining {.smaller}

When joining Points (starbucks) to Polygons (counties):

```{r}
#| eval: false
# Join points to polygons (many points per polygon)
points_in_polys <- st_join(polygons, points)

# Aggregate: count, mean, max, list(), etc.
polygon_summaries <- points_in_poly |>
  group_by(polygon_id, polygon_name) |>
  summarise(
    count = n(),
    mean_value = mean(value, na.rm = T),
    types = list(unique(type)),
    .groups = "drop"
  )
```

Result: 1 row per polygon with computed summaries.

# Comprehensive Water Resources Example

## Multi-Source Watershed Analysis {.smaller}

Combine stream gauges, watersheds, and infrastructure using spatial relationships:

```{r}
#| eval: false
# 1️⃣ Which gauges are in which watershed?
gauges_watersheds <- st_join(gauges, watersheds, join = st_within)

# 2️⃣ Which dams are near gauges? (within 10 km downstream)
dams_near_gauges <- st_filter(
  dams, 
  gauges,
  .predicate = st_is_within_distance,
  dist = 10000
)

# 3️⃣ Summary: Count gauges, mean discharge, dam presence per watershed
summary <- gauges_watersheds |>
  mutate(has_dam = huc_id %in% (dams_near_gauges$huc_id)) |>
  group_by(huc_id, watershed_name, has_dam) |>
  summarise(
    n_gauges = n(),
    mean_discharge_m3s = mean(discharge, na.rm = T),
    max_discharge_m3s = max(discharge, na.rm = T),
    .groups = "drop"
  ) |>
  arrange(desc(mean_discharge_m3s))
```

**Output**: Watershed summary with gauge counts, discharge statistics, and dam presence—ready to investigate relationships between infrastructure and streamflow patterns.

---

This layered approach (join, filter, mutate, group, summarise) is the template for most spatial analyses.

# Clipping & Spatial Subsetting

## Filtering vs. Clipping

So far we've **filtered** and **joined**—neither changes the geometry itself.

**Clipping** is different: it *modifies* geometry by intersecting it with another geometry.

- **Filter**: Select entire features where a spatial relationship is TRUE
- **Clip**: Trim the geometry of features to fit within a boundary; return new boundaries

Clipping only works with Lines and Polygons (not Points—they either fall in or out).

## Spatial Subsetting with `[,]`

The default data frame subsetting operator `[,]` implements spatial intersection when used with `sf` objects:

```{r}
#| eval: false
# All Starbucks in Washington state
wa_starbucks <- starbucks[wa, ]
```

This is equivalent to:

```{r}
#| eval: false
st_intersection(starbucks, wa)
```

Result: Returns geometries that fall within (or on the boundary of) `wa`.

## Clipping Lines and Polygons {.smaller}

When you clip a **line** or **polygon** to a boundary, the geometry is **trimmed**:

```{r}
#| eval: false
# Clip CONUS boundary to state borders
state_boundaries <- st_intersection(conus_boundary, state_borders)

# Clip stream network to a watershed
watershed_streams <- st_intersection(nhdplus_flowlines, watershed_boundary)
```

**Output**: New geometries with coordinates moved to the intersection boundary.

. . .

::: callout-important
Clipping is computationally expensive for large geometries. Simplify first if possible (especially if clipping boundaries have thousands of vertices).
:::

## Clipping Visualization {.smaller}

```{r}
#| eval: true
#| echo: false
#| fig.align: center
#| fig.width: 12
#| fig.height: 5

# Original: All cities
p_before <- ggplot(states) + 
  geom_sf(fill = "white", color = "black", linewidth = 0.3) + 
  geom_sf(data = wa, fill = "steelblue", alpha = 0.2) +
  geom_sf(data = cities, color = "red", size = 2) + 
  theme_void() +
  labs(title = paste0("All cities (", nrow(cities), " total)"))

# After clipping to WA
cities_in_wa <- st_intersection(cities, wa)

p_after <- ggplot(wa) + 
  geom_sf(fill = "steelblue", alpha = 0.2, color = "black", linewidth = 0.5) + 
  geom_sf(data = cities_in_wa, color = "red", size = 2) + 
  theme_void() +
  labs(title = paste0("Cities clipped to WA (", nrow(cities_in_wa), " kept)"))

gridExtra::grid.arrange(p_before, p_after, nrow = 1)
```

Left: All cities in region. Right: Only cities *within* WA boundary (geometry trimmed at edge).

## Why Simplify?

As geometries become more complex, operations slow down. Simplification reduces vertices while preserving shape.

**Reasons to simplify**:
- Reduce computation time for clipping, unions, and spatial operations
- Decrease file size and network bandwidth for data transfer
- Match the resolution appropriate to your analysis scale
- Remove noise from digitized boundaries

**Trade-off**: Detail vs. speed. A map at national scale needs less detail than a county-scale map.

## Douglas-Peucker Algorithm

The industry standard for simplification:

1. **Mark** first and last points as kept
2. **Find** the point farthest from the first-last line segment
3. **If** that point is closer than `tolerance` units → discard everything between first and last
4. **Else** → mark it as kept and recursively simplify subsections

Result: Removes ~70-90% of vertices while preserving overall shape.

## `st_simplify()` in R

`sf` provides `st_simplify()` using GEOS (Douglas-Peucker):

```{r}
#| eval: false
# Simplify with 10 km tolerance (in CRS units)
usa_simple <- st_simplify(usa, dTolerance = 10000)

# More aggressive simplification
usa_aggressive <- st_simplify(usa, dTolerance = 100000)
```

**Trade-off**: Per-geometry basis means **topology can break** (boundaries overlap, gaps appear).

## Visvalingam-Whyatt Algorithm (Advanced)

Better than Douglas-Peucker; preserves visual quality better.

**How it works**: Progressively removes points with least-perceptible area change rather than distance-based.

**In R**: Use `rmapshaper` package (wrapper around mapshaper JavaScript library):

```{r}
#| eval: false
library(rmapshaper)

# Keep 10% of points
usa_10pct <- ms_simplify(usa, keep = 0.1)

# Keep 1% of points  
usa_1pct <- ms_simplify(usa, keep = 0.01)
```

Result: Often retains 95%+ visual quality with 90%+ fewer vertices.

## Simplification: Visual Comparison {.smaller}

```{r}
#| eval: true
#| echo: false
#| fig.align: center
#| fig.width: 14
#| fig.height: 4

# Simplify with different tolerances
usa_original <- st_union(aoi_get(state = "conus"))
usa_simple_1k <- st_simplify(usa_original, dTolerance = 1000)
usa_simple_10k <- st_simplify(usa_original, dTolerance = 10000)
usa_simple_100k <- st_simplify(usa_original, dTolerance = 100000)

p1 <- ggplot(usa_original) + geom_sf() + theme_void() + 
  labs(title = paste0("Original\n(", mapview::npts(usa_original), " vertices)"),
       caption = "dTolerance = 0")

p2 <- ggplot(usa_simple_1k) + geom_sf() + theme_void() + 
  labs(title = paste0("1km tolerance\n(", mapview::npts(usa_simple_1k), " vertices)"),
       caption = "dTolerance = 1,000")

p3 <- ggplot(usa_simple_10k) + geom_sf() + theme_void() + 
  labs(title = paste0("10km tolerance\n(", mapview::npts(usa_simple_10k), " vertices)"),
       caption = "dTolerance = 10,000")

p4 <- ggplot(usa_simple_100k) + geom_sf() + theme_void() + 
  labs(title = paste0("100km tolerance\n(", mapview::npts(usa_simple_100k), " vertices)"),
       caption = "dTolerance = 100,000")

gridExtra::grid.arrange(p1, p2, p3, p4, nrow = 1)
```

Notice: As tolerance increases, vertices are removed but shape is preserved. Larger tolerances = faster operations.

## What is Tessellation?

Dividing a region into a grid of regular or irregular shapes (tiles).

**Common applications**:
- Aggregating point data (cities → grid cells)
- Organizing data for processing (satellite imagery)
- Interpolation and spatial smoothing
- Identifying spatial patterns (MAUP—Modifiable Areal Unit Problem)

## Tessellation: Square vs. Hexagonal vs. Voronoi {.smaller}

```{r}
#| eval: true
#| echo: false
#| fig.align: center
#| fig.width: 15
#| fig.height: 5

# Create sample region
region <- st_union(aoi_get(state = c("WA", "OR")))

# Create different tessellations
sq_grid <- st_make_grid(region, n = c(8, 10)) |>
  st_as_sf() |> mutate(id = 1:n())

hex_grid <- st_make_grid(region, n = c(8, 10), square = FALSE) |>
  st_as_sf() |> mutate(id = 1:n())

# For Voronoi, create sample points
set.seed(123)
sample_pts <- st_sample(region, size = 30)
voronoi <- st_voronoi(st_union(sample_pts)) |> 
  st_cast() |> st_as_sf() |> mutate(id = 1:n()) |>
  st_intersection(st_union(region))

p_sq <- ggplot(sq_grid) + geom_sf(fill = NA, color = "navy") + 
  geom_sf(data = region, fill = NA, color = "red", linewidth = 1) +
  theme_void() + labs(title = "Square Grid")

p_hex <- ggplot(hex_grid) + geom_sf(fill = NA, color = "navy") + 
  geom_sf(data = region, fill = NA, color = "red", linewidth = 1) +
  theme_void() + labs(title = "Hexagonal Grid")

p_vor <- ggplot(voronoi) + geom_sf(fill = NA, color = "navy") + 
  geom_sf(data = sample_pts, color = "red", size = 2) +
  geom_sf(data = region, fill = NA, color = "darkred", linewidth = 1) +
  theme_void() + labs(title = "Voronoi (red = anchor points)")

gridExtra::grid.arrange(p_sq, p_hex, p_vor, nrow = 1)
```

**Square**: Simple, regular. **Hexagon**: Better edge effects, equal neighbors. **Voronoi**: Based on anchor points, useful for interpolation.

## Creating Tessellations: The Pattern {.smaller}

All tessellation functions (`st_voronoi()`, `st_triangulate()`, `st_make_grid()`) return geometry collections. The workflow is consistent:

```{r}
#| eval: false
# 1. Create tessellation
tess_raw <- st_voronoi(st_union(points))

# 2. Cast to simplify (review: Week 2 - Feature vs. Set level)
tess_clean <- st_cast(tess_raw)

# 3. Convert to sf object to add attributes
tess_sf <- tess_clean |>
  st_as_sf() |>
  mutate(id = 1:n())
```

**Why cast?** Tessellations return complex nested structures. `st_cast()` flattens and validates them, preventing topology errors in downstream operations.

Partition space based on **anchor points**—each polygon is the area closest to its anchor:

```{r}
#| eval: false
# Convert points to multipoint, create Voronoi
voronoi_grid <- st_voronoi(st_union(anchor_points)) |>
  st_cast() |>
  st_as_sf() |>
  mutate(id = 1:n())
```

**Useful for**:
- Nearest-neighbor search
- Facility location optimization
- Spatial interpolation (e.g., rainfall from gauges)
- Largest empty area problems

## Delaunay Triangulation: Visual Example {.smaller}

```{r}
#| eval: true
#| echo: false
#| fig.align: center
#| fig.width: 12
#| fig.height: 5

set.seed(456)
pts <- st_sample(region, size = 25)

# Delaunay triangulation
delaunay <- st_triangulate(st_union(pts)) |>
  st_cast() |> st_as_sf() |>
  st_intersection(st_union(region))

# Voronoi for comparison
voronoi2 <- st_voronoi(st_union(pts)) |> 
  st_cast() |> st_as_sf() |>
  st_intersection(st_union(region))

p_del <- ggplot(delaunay) + geom_sf(fill = NA, color = "navy", linewidth = 0.5) + 
  geom_sf(data = pts, color = "red", size = 2) +
  theme_void() + labs(title = "Delaunay Triangulation")

p_vor2 <- ggplot(voronoi2) + geom_sf(fill = NA, color = "darkgreen", linewidth = 0.5) + 
  geom_sf(data = pts, color = "red", size = 2) +
  theme_void() + labs(title = "Voronoi (Dual of Delaunay)")

gridExtra::grid.arrange(p_del, p_vor2, nrow = 1)
```

**Delaunay**: Triangulates anchor points. **Voronoi**: Complements it—circumcenters of Delaunay triangles are Voronoi vertices.

Connect anchor points with triangles such that no point lies inside any triangle's circumcircle:

```{r}
#| eval: false
triangles <- st_triangulate(st_union(anchor_points)) |>
  st_cast() |>
  st_as_sf() |>
  mutate(id = 1:n())
```

**Relationship**: Delaunay triangles are the *dual* of Voronoi polygons (circumcenters of triangles = Voronoi vertices).

**Used in**:
- Terrain modeling (TINs—Triangular Irregular Networks)
- Flood mapping
- 3D surface interpolation

## Modifiable Areal Unit Problem (MAUP)

**The problem**: Aggregate your data differently, get different results.

**Why it matters**: Summary statistics (mean, proportion) depend on:
- **Scale** of aggregation units (county vs. state)
- **Shape** of units (regular vs. irregular)

**Example**: Gerrymandering—redraw voting districts, change political outcomes.

::: callout-warning
Be cautious when choosing aggregation units. Different zonation schemes produce different spatial patterns. Always report the units used.
:::



Be cautious when choosing aggregation units. Different zonation schemes produce different spatial patterns. Always report the units used.
:::

# Putting It All Together: Typical Lab Workflow

## The Complete Pattern {.smaller}

Here's the workflow you'll follow in Lab 3 (and most spatial analyses):

```r
# 1️⃣ Read and prepare data
dams <- read_csv("data.csv") |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326) |>
  st_transform(5070)  # Project

# 2️⃣ Create tessellation
boundary <- st_union(aoi_get(...))
grid <- st_make_grid(boundary, n = c(70, 50)) |>
  st_as_sf() |>
  mutate(id = 1:n())

# 3️⃣ Simplify boundary if needed (before clipping!)
boundary_simple <- ms_simplify(boundary, keep = 0.05)

# 4️⃣ Crop tessellation to boundary
grid_cropped <- st_intersection(grid, boundary_simple)

# 5️⃣ Point-in-polygon: Count dams per grid cell
dams_by_grid <- st_join(grid_cropped, dams) |>
  group_by(id) |>
  summarise(n = n())  # n = count of dams per cell

# 6️⃣ Visualize with conditional highlighting
ggplot(dams_by_grid) +
  geom_sf(aes(fill = n), color = NA) +
  scale_fill_viridis_c() +
  gghighlight(n > mean(n) + sd(n))  # Highlight high-count cells
```

. . .

**Key decisions to make**:
- Which tessellation type? (grid, hex, Voronoi)
- How much to simplify? (keep = 0.1 or 0.05?)
- Which predicate for joining? (st_within, st_intersects)
- How to aggregate? (count, mean, sum, list)

## Why Each Step Matters {.smaller}

| Step | Why | Function |
|---|---|---|
| **Project early** | All spatial ops in same CRS | `st_transform()` |
| **Simplify before clipping** | Large boundaries = slow intersections | `ms_simplify()` |
| **Cast after tessellation** | Avoids topology errors | `st_cast()` |
| **Join with correct predicate** | Changes what counts as "in" | `st_within()` vs `st_intersects()` |
| **Group then count** | Many-to-many joins need aggregation | `group_by() + summarise()` |
| **Conditional viz** | MAUP: different patterns for different scales | `gghighlight()` |

# Summary

## This Week: Topology, Operations & Tessellation

You now know:

- **Interior, Boundary, Exterior** — How geometry is structured
- **DE-9IM matrix & `st_relate()`** — Mathematical foundation and full topological codes
- **Six named predicates** — Intersects, Touches, Within, Contains, Overlaps, Crosses
- **Spatial filtering** — `st_filter()` subsets data by relationship
- **Spatial joining** — `st_join()` combines datasets; watch for many-to-many matches
- **Clipping** — `st_intersection()` modifies geometry to fit boundaries
- **Simplification** — Douglas-Peucker & Visvalingam algorithms reduce vertices
- **Tessellation** — Square grids, hexagons, Voronoi, Delaunay triangulation
- **MAUP** — How aggregation unit choice affects analysis results

. . .

**Core principle**: Spatial problems are relational problems solved through topology, set operations, and spatial aggregation.

---

**Next week**: Raster data—gridded coverage model (pixels, not polygons).

**Key takeaway**: Choose your tessellation and aggregation scheme carefully. It shapes your results.
