---
title: 'Lab 2: Distances and Projections'
subtitle: 'Hints & Tricks'
format:
  html:
    toc: true
    include-before-body: ../../slides/header.html
    include-after-body:  ../../slides/footer-annotations.html
---

```{r, echo = FALSE}
col_dist <- RColorBrewer::brewer.pal(11, "RdGy")
```

```{r, eval = TRUE, message = F, warning = F, echo  = FALSE}
# SPDS
library(tidyverse)
library(sf)
library(units)

# Data
library(USAboundaries)
library(rnaturalearth)

# Visualization
library(gghighlight)
library(ggrepel)
library(knitr)
```

```{r, echo = FALSE, message = FALSE, warning=FALSE}
cities = read_csv("../data/uscities.csv") |> 
  st_as_sf(coords = c("lng", "lat"), crs = 4326) |> 
  st_transform(5070) |> 
  filter(!state_name %in% c("Alaska", "Hawaii", "Puerto Rico"))
```

# Question 1:

### Making Spatial Objects & Coordinate Transformation

Spatial objects (sf) can be built from a vector of X and Y values in addition to a coordinate reference system (CRS). For example:

```{r}
df <- data.frame(name = state.name, 
                X = state.center$x, 
                Y = state.center$y)
head(df)

# Geographic Coordinate System (GCS)
(df_sf_gcs = st_as_sf(df, 
                      coords = c("X", "Y"), 
                      crs = 4269))

ggplot() + 
  geom_sf(data = df_sf_gcs) + 
  coord_sf(datum = st_crs(df_sf_gcs)) +
  theme_linedraw()

# Projected Coordinate System (PCS)
# st_transforms converts from one reference system to another
(df_sf_pcs = st_transform(df_sf_gcs, 5070))

ggplot() + 
  geom_sf(data = df_sf_pcs) + 
  coord_sf(datum = st_crs(df_sf_pcs)) +
  theme_linedraw()
```

# Question 2: 

### `st_distance` review

```{r}
# Three most populous cities in the USA
(big3 = cities |> 
   select(city, population) |> 
   slice_max(population, n = 3))

# Fort Collins
(foco = filter(cities, city == "Fort Collins") |> 
    select(city, population))

# Distance from foco to population centers
st_distance(big3, foco)
```

There are two notable things about this result:

1. It has `units`
2. It is returned as a matrix, even though foco only had one row

This second point highlights a useful feature of `st_distance`, namley, its ability to return distance matrices between all combinations of features in `x` and `y`.

### `units` review

```{r, error = FALSE}
st_crs(big3)$units
```

Units can be converted using `units::set_units`. For example, 'm' can be converted to 'km'.

```{r}
big3 = mutate(big3, 
              dist_to_foco = st_distance(big3, foco),
              dist_to_foco = set_units(dist_to_foco, "km")) 

(big3$dist_to_foco)
```

If needed, drop units with `units::drop_units()` before numeric operations.

### `Geometry` review

`st_combine`, `st_union`, and `st_cast` are commonly used together to build line boundaries for distance calculations. Use `st_union()` to dissolve internal boundaries for an outer border and `st_combine()` to preserve internal state lines.

```
# Example: cast to MULTILINESTRING for border distance calculations
# line_border = st_cast(st_union(states), "MULTILINESTRING")
```

# Question 3:

Examples of `gghighlight` and `ggrepel` usage are included below; use `stat = "sf_coordinates"` when labeling `sf` points.

```
# Example mapping pattern:
# ggplot() + geom_sf(data = states) + geom_sf(data = cities, aes(color = dist_to_border)) + 
# ggrepel::geom_label_repel(data = top_cities, aes(label = city, geometry = geometry), stat = "sf_coordinates")
```

***

These hints are intentionally practical; keep them as optional help for students who are stuck on geometry/CRS debugging and plotting mechanics.
